On joint estimation of ancestral states and phylogenetic model parameters

### 2017-08-01 joint_inf.tex
Mostly just cleaning up; new title :D

\title{Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States}
\author{Shaw, Matsen, Minin}

@@

%EM This sentence should be made more precise-- it certainly changes the value of the log likelihood-- you mean this restriction does not change the ML.
Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
%EM complete to introduce: We now note that exchanging $x_1$ and $x_2$, or $y_1$ or $y_2$, does not change ...
%EM please be a little more specific than just $ell$, because there are lots of variants-- e.g. likelihood with what data?
Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities on the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Consider only the Farris-tree topology.
Assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log((1+w'))
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
    &\qquad + p_{13}\cdot\log((1-w')).
\end{align*}
When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
The estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.

@@

Since we do not know the true values of $x$ and $y$, standard methods will not be in the situation in Figure~\ref{fig:w_hat}.
The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}=x$ and $\hat{y}=y$.
If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.

@@

We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
Even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.

>>>>

\title{On joint estimation of ancestral states and phylogenetic model parameters}
\author{Shaw, Matsen, Minin}

@@

Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
We now show that, in the case of the Farris tree topology, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that fixing both the top two branch lengths to be equal and the bottom two branch lengths to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch lengths.
Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities on the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Consider only the Farris-tree topology.
We will consider two settings, corresponding to whether some branch lengths are known ahead of time or not.

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log((1+w'))
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
    &\qquad + p_{13}\cdot\log((1-w')).
\end{align*}
When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
We find that the estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.

@@

Second, we consider the more standard case where the true values of $x$ and $y$ are not known.
The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}=x$ and $\hat{y}=y$.
If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.

@@

We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
In summary, even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.
